---
title: "Choroplet Maps with `leaflet` and `ggplot`"
description: |
  Plotting UCR violent crime with choroplet maps using different packages
date: 11-15-2020
categories:
  - map
  - choroplet
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
library(dplyr)
library(tigris)
library(leaflet)
library(RColorBrewer)
library(ggplot2)
library(viridis)
```

## Data preperation

UCR data can be downloaded from [ICPSR](https://www.icpsr.umich.edu/web/pages/index.html) web page as a `.rda` file. Let's read the data into R and rename it:

```{r load_data, message=FALSE, warning=FALSE}
load("../../datasets/37059-0001-Data.rda")
ucrRaw <- da37059.0001
```

We will merge UCR dataset with geographic information data. We need to create a common key:

```{r fips_code}
for(i in 1:length(ucrRaw$FIPS_ST)){
  if(ucrRaw$FIPS_ST[i] < 10){
    ucrRaw$CHR_FIPS_ST[i] <- paste(0, as.character(ucrRaw$FIPS_ST[i]), sep = "")
  }else{
    ucrRaw$CHR_FIPS_ST[i] <- as.character(ucrRaw$FIPS_ST[i])
  }
}
```

We need to calculate state level crime data and take out US territories outside mainland:

```{r state_level}
ucrStates <- ucrRaw %>% 
  select(c(7:8,10:57)) %>% 
  group_by(CHR_FIPS_ST) %>% 
  summarise(across(c(1:2,4:49), sum), COVIND = mean(COVIND, na.rm = TRUE)) %>% 
  filter(CHR_FIPS_ST != "02" , CHR_FIPS_ST != "78" , CHR_FIPS_ST != "15" , CHR_FIPS_ST != "60" , 
         CHR_FIPS_ST != "66" , CHR_FIPS_ST != "69" , CHR_FIPS_ST != "72")
```

There are different ways to incorporate geographic information onto the analysis. Here we use `tigris` package:

```{r geodata, include=FALSE}
states <- states(cb=T) %>% 
  filter(GEOID != "02" , GEOID != "78" , GEOID != "15" , GEOID != "60" , 
           GEOID != "66" , GEOID != "69" , GEOID != "72" )
```

Finally we merge crime data with geographic information data and calculate a new variable showing violent crime per 1,000 people for each state:

```{r geodata1}
shp_ucrStates <- geo_join(states, ucrStates, "GEOID", "CHR_FIPS_ST")
shp_ucrStates$VCper1000 <- (shp_ucrStates$P1VLNT / shp_ucrStates$CPOPARST) * 1000
```

## `leaflet` Map
Leaflet is a powerful package for producing interactive maps:

```{r leaflet_map}
m <- leaflet() %>% 
  addProviderTiles("CartoDB.Positron") %>%
  setView(-98.483330, 38.712046, zoom = 4)

bins <- c(0, 0.5, 1, 1.5, 2, 2.5, 3, 3.5, Inf)
pal <- colorBin("YlOrRd", domain = shp_ucrStates$VCper1000)
m %>% addPolygons(data = states,
  fillColor = ~pal(shp_ucrStates$VCper1000),
  weight = 2,
  opacity = 1,
  color = "white",
  dashArray = "3",
  fillOpacity = 0.7,
  highlight = highlightOptions(
    weight = 5,
    color = "#666",
    fillOpacity = 0.7,
    bringToFront = TRUE),
  label = ~paste0(NAME, ": ", formatC(shp_ucrStates$P1VLNT, format ="d", big.mark = ","))) %>% 
  addLegend(pal = pal, 
            values = shp_ucrStates$VCper1000, 
            position = "bottomright", 
            title = "Violent Crime <br />per 1000<br/>residents")
```

States are colored according to total number of reported violent crimes per 1,000 people. Hovering over the map highlights states and activate pop ups.

We can apply same logic to an individual state. Here I will draw map of New York state and color countries according to the number of homicides. First, prepare the data:

```{r include=FALSE}
#Filter NY (FIPS Code = 36) from the dataset
ny_counties <- da37059.0001 %>% 
  filter(FIPS_ST == "36")

#Convert FIPS code into character for compatibility
for(i in 1:length(ny_counties$FIPS_CTY)){
  if(ny_counties$FIPS_CTY[i] < 10){
    ny_counties$CHR_FIPS_CTY[i] <- paste("00", as.character(ny_counties$FIPS_CTY[i]), sep = "")
  }else if(ny_counties$FIPS_CTY[i] < 100 & ny_counties$FIPS_CTY[i] > 9 ){
      ny_counties$CHR_FIPS_CTY[i] <- paste("0", as.character(ny_counties$FIPS_CTY[i]), sep = "")
  }else{
    ny_counties$CHR_FIPS_CTY[i] <- as.character(ny_counties$FIPS_CTY[i])
  }
}

#Download geographic data of New York State via Tigris package
geo_ny_counties <- counties(state = "NY", cb = TRUE)

#Merge geographic data with UCR data
merged_ny_counties <- geo_join(geo_ny_counties, ny_counties, "COUNTYFP", "CHR_FIPS_CTY")
```

Now we are ready to draw the map. Code is similar to the previous one. I changed latitude and longtitude, zoom level, bin width and variable name:

```{r}
m <- leaflet() %>% 
  addProviderTiles("CartoDB.Positron") %>%
  setView(-75.59655, 42.921241, zoom = 6)

bins <- c(0, 1, 5, 7, 10, 15, 25, 180)
pal <- colorBin("YlOrRd", domain = merged_ny_counties$MURDER, bins = bins)
m %>% addPolygons(data = merged_ny_counties,
                  fillColor = ~pal(merged_ny_counties$MURDER),
                  weight = 2,
                  opacity = 1,
                  color = "white",
                  dashArray = "3",
                  fillOpacity = 0.7,
                  highlight = highlightOptions(
                    weight = 5,
                    color = "#666",
                    fillOpacity = 0.7,
                    bringToFront = TRUE),
                  label = ~paste0(NAME, ": ", formatC(merged_ny_counties$MURDER, format ="d", big.mark = ","))) %>% 
  addLegend(pal = pal, values = merged_ny_counties, position = "bottomright", title = "Homicides")
```


## `ggplot` map
We can produce static and publication ready maps with `ggplot`:

```{r ggplot_map}
shp_ucrStates %>% 
  ggplot(aes(fill = VCper1000, color = VCper1000)) +
  geom_sf() +
  coord_sf(crs = 5070, datum = NA) +
  scale_fill_viridis(direction = -1) + 
  scale_color_viridis(direction = -1) +
  labs(title = "US violent crime map",
       subtitle = "Number of violent crimes per 1,000 residents reported by LE agencies", 
       caption = "Data Source: UCR - Unified Crime Reports ; 2019",
       fill = "Violent Crimes") + guides(color=FALSE)
```

Again we can do a similar map for individual states:

```{r}
merged_ny_counties %>% 
  ggplot(aes(fill = MURDER, color = MURDER)) +
  geom_sf() +
  coord_sf(crs = 5070, datum = NA) +
  scale_fill_viridis(direction = -1) + 
  scale_color_viridis(direction = -1) +
  labs(title = "New York State",
       subtitle = "Number of homicides reported by LE agencies", 
       caption = "Data Source: UCR - Unified Crime Reports ; 2019",
       fill = "Homicides") + guides(color=FALSE)
```
